---
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r preamble2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Polar maps

What is required for mapping in polar regions? 

* shape-feature orientations, the dateline and the poles
* handling of data in long-lat, understanding of metrics and shape (area, length, angle)
* projected maps can be very hard to do, each has its own issues
* quantities like vector fields
* conventions for global data -180/180, 0/360





## Mapping in R

```{r setup-02}
library(sp)
library(raster)
library(maps)
library(rgdal)
data("wrld_simpl", package = "maptools")
```


The oldest general mapping tool in R is the `maps` package. It has a simple whole-world coastline data set for immediate use. 

```{r maps-package}
maps::map()
```

The data underlying this live map is available by capturing the output as an
actual object. 


If we look carefully at the southern edge and the eastern edge, notice that the
coastline for Antarctica does not extend to the south pole, and the Chukotka
region of Russia east of 180 longitude is not in the western part of the map.


```{r maps-assign}
maps_c <- maps::map(plot = FALSE)

lonlat <- cbind(maps_c$x, maps_c$y)

plot(lonlat, pch = "+", cex = 0.4, axes = FALSE)
lines(lonlat, col = "dodgerblue")

abline(h = c(-90, 90), v = c(-180, 180))

```

A very similar and slightly more modern data set is available in the `maptools` package.


(It's not sensible use `maps` or `maptools` for coasline or country boundary data, but they
are handy for exploring concepts ). 

```{r wrld-simpl}
data("wrld_simpl", package = "maptools")

plot(wrld_simpl)
```

This data set aligns exactly to the  conventional -180/180 -90/90 extent of the longitude/latitude projection.

```{r lon180-lat90}
plot(0, type = "n", axes = FALSE, xlab = "", ylab = "", xlim = c(-180, 180), ylim = c(-90, 90))
rect(xleft = -180, ybottom = -90, xright = 180, ytop = 90, border = "darkred", lwd = 4, lty = 2)
plot(wrld_simpl, add = TRUE)
```

## Exercises 01

1. How can we find the longitude and latitude ranges of the maps data `maps_c` and the maptools data `wrld_simpl`? 
2. Can we draw polygons with a fill colour with the maps package? 


Answer 1: `range(m$x, na.rm = TRUE)` `range(m$y, na.rm = TRUE`) also `m$range`

Answer 2: `polygon(lonlat, col = "grey")` does not work, and `map(mp, fill = TRUE, col = "grey")` does not work, but `maps::map(fill = TRUE, col = "grey")` does seem to work.


What's going on? Look at the very south-eastern corner of the map. The "coastline" has been extended to the very south boundary of the available area. 

```{r south-east}
plot(0, type = "n", axes = FALSE, xlab = "", ylab = "", xlim = c(-150, 180), ylim = c(-90, -60))
plot(wrld_simpl, add = TRUE, col = "grey")
rect(xleft = -180, ybottom = -90, xright = 180, ytop = 90, border = "darkred", lwd = 4, lty = 2)

maps::map(add = TRUE, col = "dodgerblue", lwd = 3)
```

When we add the old maps coastline see that it does not extend to 90S and it does not traverse the southern boundary.

One reason for this is that if we choose a projection where the east and west edges of the Antarctic coastline meet then we get what looks a fairly clean join.

```{r laea}
## scale factor
f <- 3e6
plot(rgdal::project(lonlat, "+proj=laea +lat_0=-90 +datum=WGS84"), asp = 1, type = "l",
     xlim = c(-1, 1) * f, ylim = c(-1, 1) * f, xlab = "", ylab = "")

```

If we try the same with `wrld_simpl` it's not as neat.  We have a strange "seam" that points exactly to the south pole (our projection is centred on longitude = 0, and latitude = -90. 



```{r wrld}
plot(sp::spTransform(wrld_simpl, "+proj=laea +lat_0=-90 +datum=WGS84"), asp = 1,
     xlim = c(-1, 1) * f, ylim = c(-1, 1) * f, xlab = "", ylab = "", lwd = 3)
abline(v = 0, h = 0, lty = 2, col = "grey")
```


## Let's use the maps data!

In `maps_c` we have the maps data structure, and this looks promising.

```{r maps-data}
str(maps_c)

mp <- maps_c
pxy <- rgdal::project(lonlat, "+proj=laea +lat_0=-90 +datum=WGS84")
mp$x <- pxy[,1]
mp$y <- pxy[,2]
mp$range <- c(range(mp$x,na.rm = TRUE), range(mp$y, na.rm = TRUE))
mp$range
plot(c(-1, 1) * f, c(-1, 1) * f, type = "n", asp = 1)
maps::map(mp, add = TRUE)


## but it doesn't take much to go awry
plot(c(-1, 1) * f, c(-1, 1) * f, type = "n", asp = 1)
maps::map(mp, add = TRUE, fill = TRUE, col = "grey")

```

The problem is that the maps database has enough internal structure to join lines correctly, with `NA` gaps between different connected linestrings, but not enough to draw these things as polygons. A similar problem occurs in the default projection. While `wrld_simpl` has been extend by placing two dummy coordinates at the east and west versions of the south pole, this data set does not have those.

We have to look quite carefully to understand what is happening, but this is wrapping around overlapping itself and so close to the southern bound we barely notice.


```{r no-pole}
plot(0, type = "n", axes = FALSE, xlab = "", ylab = "", xlim = c(-180, -110), ylim = c(-90, -60))

rect(xleft = -180, ybottom = -90, xright = 180, ytop = 90, border = "darkred", lwd = 4, lty = 2)

maps::map(add = TRUE,col = "grey", fill = TRUE)

maps::map(col = "grey", fill = TRUE)


mpmerc <- maps_c
pxy <- rgdal::project(lonlat, "+proj=merc +datum=WGS84")
mpmerc$x <- pxy[,1]
mpmerc$y <- pxy[,2]
mpmerc$range <- c(range(mpmerc$x,na.rm = TRUE), range(mpmerc$y, na.rm = TRUE))
mpmerc$range

## the catastrophe made a little clearer 
plot(0, xlim = range(mpmerc$range[1:2]), ylim = c(mpmerc$range[1], 0))
maps::map(mpmerc,  fill = TRUE, col = "grey", add = TRUE)


```



## Reprojecting to polar regions

In essence it should be easy, but details really matter. 

```{r raster-reprojection}
sst <- raster::raster(system.file("nc/reduced.nc", package = "stars", mustWork = TRUE))

plot(raster::projectRaster(sst, crs = "+proj=laea +lat_0=-90"))
plot(raster::projectRaster(rotate(sst), crs = "+proj=laea +lat_0=-90"))
```

Loxodromes need densifying, and not along great circles. 

The pole is undefined in Mercator.

Sensible polar projections are stereographic and Lambert Azimuthal Equal Area. These are more or less identical over a large area, so knowing what is in use really matters!


```{r polar-laea}
data("wrld_simpl", package = "maptools")
laea <- "+proj=laea +lon_0=0 +lat_0=-90 +datum=WGS84"
stere <- "+proj=stere +lon_0=0 +lat_0=-90 +datum=WGS84"

wm <- wrld_simpl[coordinates(wrld_simpl)[,2] < 30, ]
 
## remember that plotting in R is usually *NOT* coordinate system aware ...

plot(spTransform(wm, laea))
plot(spTransform(wm, stere), add = TRUE, border = "firebrick")
points(0, 0, cex = 40)
```


This video explains the Lambert Azimuthal Equal Area projection: https://www.youtube.com/watch?v=quzIU4nL9ig  (crux at 2:25 with the spherical shells). 

## Demonstration of the projection concept

rgl plots Stereographic, LAEA, Orthographic





