
# Getting Set Up {#setup}

We need the following packages installed, using `remotes` is the best way to install as it will not re-install any package if it's already up to date. 

```{r setup, eval = FALSE}
## make sure we have the packages we need
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
req1 <- c("ggplot2", "maps", "maptools", "raster", "sp", "dplyr", "rgdal")
remotes::install_cran(req1)
```


## Data sets

Obtain some extra data sets that will be used in this session. 

```{r extra-data, eval=FALSE}
if (!file.exists("polar-mapping.Rdata")) {
 download.file("https://github.com/mdsumner/polar-mapping-oghub/raw/master/data-raw/polar-mapping.Rdata", 
               
               "polar-mapping.Rdata", mode = "wb")
}
load("polar-mapping.Rdata")
```


## Alternative 1

If you have problems and can't get R set up as above, try this Binder badge - just click it and you should get a ready-to-go Rstudio session where the code already works. 

[![Launch Rstudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/mdsumner/polar-mapping-oghub/master?urlpath=rstudio)

(It shouldn't take more than a few minutes to launch, but note that a session will only last for 12 hours or so and you will need to keep active with the session, it will time out after 10min or so). 

## Alternative 2

If you get really stuck, try https://rstudio.cloud/

There you should be able to run the code above with no problem. 



