---
output: html_document
editor_options: 
  chunk_output_type: console
---


# SOmap

The `SOmap` package is intended to solve some of these problems, and provide an easier way to produce nice-looking maps of Antarctica and the Southern Ocean. It is primarily focused on maps in polar stereographic projection (although the `SOmap_auto` function extends this to other projections). `SOmap` won't necessarily get you exactly the map you want. But if it doesn't, it should hopefully get you close enough that you can make modifications to suit your exact purposes.

Please bear in mind that `SOmap` is still in development, and so its functionality (function parameters and/or behaviour) may change.

By default, `SOmap` works with base graphics (and associated functionality from packages such as `raster` and `sp`). It is also possible to work with `ggplot2`-based graphics, as described below.

Start by loading the `SOmap` package:

```{r somap_init, cache = FALSE, eval=FALSE}
library(SOmap)
## also define a colour map to use for some examples
my_cmap <- colorRampPalette(c("#4D4140", "#596F7E", "#168B98",
                              "#ED5B67", "#E27766", "#DAAD50", "#EAC3A6"))(51)
```

```{r somap_pkgs, echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, eval=FALSE}
## ensure these are loaded, library calls below are cached and might not get evaluated every run
library(sp)
library(raster)
library(ggplot2)
```

DRAFT, this might be illustrated or not


## Circumpolar maps

A basic circumpolar map in polar stereographic projection:

```{r somap1, eval=FALSE}
SOmap()
```

`SOmanagement()` provides a number of contextual layers such as MPA boundaries and management zones.

```{r somap2, eval=FALSE}
SOmap(trim = -40) ## plot to 40S
## add the exclusive economic zones management layer
SOmanagement(eez = TRUE)
```

## Adding points

```{r somap_pts1, eval=FALSE}
## some longitude/latitude data
library(sp)
my_points_ll <- data.frame(lon = seq(0, 350, by = 10), lat = -55, z = runif(36))
coordinates(my_points_ll) <- c("lon", "lat")
projection(my_points_ll) <- "+proj=longlat +datum=WGS84"
```

This needs to be reprojected to match our map before plotting. The `SOproj` function does this:

```{r somap_pts2, eval=FALSE}
## reproject to our SOmap projection
my_points <- SOproj(my_points_ll)
## and plot
SOmap()
plot(my_points, col = "blue", add = TRUE)
```

Or use `SOplot` to reproject and plot in one step:

```{r somap_pts3, eval=FALSE}
SOmap()
SOplot(my_points_ll, col = "blue")
```
